Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo

### 2016-01-18 notes/reviews/response_to_reviews.tex

%opening
\title{Bayesian Learning of Degenerate Linear~Gaussian~State~Space~Models using~Markov~Chain~Monte~Carlo \\ Response to Reviews}
\date{January 2016}

\begin{document}

\maketitle

Thank you for taking the time to review our paper. We are sure it is a better paper as a result of your scrutiny and the consequent changes. Here is our responses to the points raised.

@@

\begin{response}
 A good point. We have added algorithm pseudo-code describing the steps of the complete MCMC kernel. It was an oversight to leave this out.
\end{response}

\begin{review}
In Section III-C, the authors propose to use a collapsed Gibbs move to more efficiently exploit the target on the space of $F$ and $Q$, alternatively. I think the authors should discuss in more details why such collapsed Gibbs is important in the text. Moreover, the importance of such steps could be easily illustrated in the results section so the reader can appreciate the benefit of using such a strategy.
\end{review}

\begin{response}
 There is a misunderstanding here, probably due to the confusing terminology. The ``collapsed Gibbs move'' simply means the composition of first targeting $\pi(F, Q)$ with Metropolis-Hastings followed by a draw from $\pi(x_{1:T}|F, Q)$. These moves are essential for the sampler to be irreducible, and they are woefully inefficient. To speed things up, we also include exact draws from the the parameter conditional $\pi(F, Q|x_{1:T})$. These are inspired by the Gibbs sampler for the full rank algorithm, but cannot fully explore the parameter space, so we refer to them as constrained Gibbs moves.

 It does not make sense to run the sampler without the Metropolis-Hastings moves. It will be stuck in whatever subspace we happen to initialise it in. Similarly, the constrained parameter conditional draws are necessary for sampling the covariance eigenvalues. In short, there is no step we could leave out without invalidating the sampler.

@@

 We have substantially re-arranged section III and added further explanation of these concepts. We hope it is clearer now.
\end{response}

\begin{review}
My main concern is regarding the simulation section.

Firstly, the comments are just a brief summary of results from the figures without any comments or discussions on why such results are obtained.
\end{review}

\begin{response}
 We have made efforts to improve the simulations section. Our aims here are:
 \begin{itemize}
  \item To show that the sampler converges and mixes satisfactorily.
  \item To show that the resulting posterior approximations are consistent with the true values.
  \item To compare using the degenerate model with the full rank model (less error but slower mixing).
  \item To demonstrate the utility of the algorithm on a practical application.
 \end{itemize}
We have added a paragraph at the start of the section explaining these aims and added some additional comments. We deliberately reserve the more general discussion of the algorithm for the final section, where we think it better belongs.
\end{response}

@@

\begin{review}
There are no figures that show the evolution of the Markov chain OR the posterior approximation of the unknown rank $r$
The authors wrote that ``the sampler identifies the true rank.... and remains at this value thereafter'' but this can be worrying - this statement could also mean that the sampler is not mixing properly for this variable $r$. The author should be more precised by saying ``Trace plots suggest that the sampler converges within a few thousand iterations. See figure 2'' - The convergence showed in this figure is only related to the variable $\xi_y$.
\end{review}

\begin{response}
We a bit zealous with trying to save space in the first draft by omitting certain figures. We have added trace plots and posterior histograms for $F$ and $Q$. We believe that the reason the sampler sits at one value of the rank throughout is simply that the posterior histogram is tightly concentrated on this value. However, as you rightly point out, this could be a sign that it is mixing poorly. We have added a second toy model in which the rank is less clearly determined by the observations, in order to demonstrate that the sampler is able to change values after convergence.
\end{response}

@@

\begin{review}
The authors could illustrate in the first toy example the benefit of the collapsed Gibbs moves by showing the performance of the sampler with vs without this strategy.
\end{review}

\begin{response}
 See earlier comments. It is not possible to omit any of the moves without the sampler becoming reducible, so this would not be a useful demonstration.
\end{response}

@@

\begin{review}
I personally think that the proposed algorithm should be illustrated with at least a different set of parameters for the toy example (e.g. different values for $\xi_y$ thus allowing to have a curve of RMSE vs $\xi_y$ )
\end{review}

\begin{response}
We do not feel that such a plot would be particularly informative, since this model is artificial and somewhat arbitrary. Inevitably, higher noise leads to larger errors... However we agree with the idea that the algorithm should be tested with some different parameter settings, which we have achieved by adding a second toy model (which also illustrates learning a posterior distribution for the rank with greater uncertainty).
\end{response}

### 2016-01-18 notes/paper-degenerate-linear-systems.tex
Improved simulations. Added second toy model.

More recently, Bayesian approaches have been developed for non-parametric learning of linear Gaussian state space models, in which full posterior distributions for the system parameters are calculated. Since this is not possible analytically (see discussion in \cite{Beal2003}), approximations must be employed. One possibility is to take a variational approach, approximating the full joint posterior distribution as a product of independent marginals \cite{Ghahramani2001,Beal2003,Barber2007}. Alternatively, the true joint posterior can be targeted using Markov chain Monte Carlo (MCMC). This has the advantage of providing consistent estimates not only of the system matrices but also of any function of these quantities (e.g. phase margins, system poles). The price we pay is in computation; it is necessary to allow the Markov chains to run until convergence in order to form good estimates. Generic Metropolis-Hastings strategies such as those adopted in \cite{Ninness2010} are liable to mix very slowly and exact too great a computational demand. However, if conjugate priors are assumed for the various system matrices, then a Gibbs sampler may be implemented which attains substantially improved efficiency and requires almost no algorithm tuning \cite{Wills2012}.

In some systems, the components of the latent state may not be free to vary independently of each other. Such systems are best modelled by a degenerate transition model, in which the transition covariance matrix is singular. This is the case, for example, when the latent state comprises the positions and velocities of multiple points on the same rigid body, meaning that their motion is deterministically coupled. \cite{Maskell2004,Gustafsson2002} consider the use of degenerate transition models for various tracking applications. The Gibbs sampler of \cite{Wills2012} is not able to handle degenerate models. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. In addition, reversible jump moves \cite{Green1995,Green2009} are introduced which enable the sampler to learn explicitly the rank of the transition covariance matrix. The new algorithm is important because it allows us to perform Bayesian learning with a broader class of linear state space models, leading to better descriptions of real systems and improved estimation performance.

@@

A basic linear state space model therefore obeys the following recursive system equations,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \tn{\ti} \\
 \ob{\ti} & = & \lgom \ls{\ti}   + \on{\ti}       ,
\end{IEEEeqnarray}
%
where $\ls{\ti} \in \reals^{\lsd}$, $\ob{\ti} \in \reals^{\obd}$. $\lgtm \in \reals^{\lsd\times\lsd}$,$\lgom \in \reals^{\obd\times\lsd}$ are the transition and observation matrices. $\tn{\ti} \in \reals^{\lsd}$ and $\on{\ti} \in \reals^{\obd}$ are the state disturbance and observation noise variables and we additionally assume that these are drawn from a Gaussian distribution,
%
\begin{IEEEeqnarray}{rCl}
 \tn{\ti} & \sim & \normalden{\tn{\ti}}{0}{\lgtv} \\
 \on{\ti} & \sim & \normalden{\on{\ti}}{0}{\lgov}     ,
\end{IEEEeqnarray}
%
where $\lgtv$ and $\lgov$ are positive definite covariance matrices. This model can be written equivalently in terms of transition and observation densities,
%
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) & = & \normalden{\ls{\ti}}{\lgtm\ls{\ti-1}}{\lgtv} \\
 \den(\ob{\ti}|\ls{\ti},\lgom,\lgov)   & = & \normalden{\ob{\ti}}{\lgom\ls{\ti}}{\lgov}      .
\end{IEEEeqnarray}
%
The initial state $\ls{1}$ may be known or may be assigned a Gaussian prior.

\subsection{Identifiability}
The system is parameterised by four matrices: $\lgtm, \lgom, \lgtv, \lgov$. In some applications, it may be appropriate to treat all of these as completely unknown. However, this scenario can be troublesome for learning algorithms, because the system is not identifiable -- we obtain the same likelihood if the latent state is multiplied by an arbitrary invertible matrix, along with corresponding multiplications of $\lgtm$, $\lgtv$ and $\lgom$. (See e.g. \cite{Roweis1999}.) To resolve the parameter ambiguity, it is necessary to constrain the values of one or more of the system matrices. One option (that taken in \cite{Roweis1999}) is to set $\lgtv$ to the identity matrix. This entails that the nature of the latent states and their relationship to the observations are entirely unknown a priori. In this paper we consider a contrasting approach, where we resolve the ambiguity by specifying the nature of the latent states and setting $\lgom$ to a pre-determined value. This approach has the advantage that the resulting model is more easily interpreted, since we can choose the latent states to be physically meaningful quantities. For example, in the motion capture application studied in section~\ref{sec:simulations}, we choose the latent states to be the positions and velocities of the markers.

@@

The solution we propose here is to use Metropolis-Hastings steps targeting the marginal to ensure irreducibility of the sampler, combined with constrained Gibbs sampling steps, which can only modify a component of the parameters, to accelerate the mixing of the chain. As in the full rank case, the target distribution will be the posterior over parameters and the state trajectory $\postden(\lgtm, \lgtv, \ls{1:\timax})$. This means that each of the Metropolis-Hastings steps targeting the marginal posterior $\postden(\lgtm, \lgtv)$ must be followed by a draw from the state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ in order to complete the iteration. The two two steps together comprise a collapsed Gibbs move \cite{Dyk2008}. Note that the forward-filtering-backward-sampling procedure can be used without modification to draw from the state conditional. See appendix~\ref{app:state-sampling}.

@@

A second decomposition will also be useful. Any positive semi-definite can be factorised in the following way,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition}
\end{IEEEeqnarray}
%
where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors with a particular structure. The transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. This factorisation is constructed using Givens rotations and thus we refer to it as the Givens decomposition. A Givens rotation matrix has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise}     ,
                                                 \end{cases}
\end{IEEEeqnarray}
%
where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions.

The Givens decomposition is important because $\tvrot$ is uniquely determined by the state trajectory, whereas $\tvfull$ is free to vary. We provide here the algorithm for conducting the Givens decomposition. The computational complexity is $\bigo{\rk\times(\lsd-\rk)}$. (Note that multiplication by a Givens matrix is $\bigo{1}$ due to sparsity.) Details and further discussion is included in appendix~\ref{app:givens-factorisation}.

\begin{algorithm}
\onehalfspacing
\small
\begin{algorithmic}[1]
 \REQUIRE Eigvenvalues and eigenvectors $\tvval,\tvvec$ of positive semi-definite matrix $\lgtv$.
 \STATE Set $\tvcross \leftarrow \idmat[\lsd\times\lsd]$, $\tvrow \leftarrow \idmat[\rk\times\rk]$, $\tvsign \leftarrow \tvvec$.
 \FOR{$i = \rk,\dots,1$}
   \FOR{$j = 1,\dots,\rk-1$}
      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{i,i})$.
      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
      \STATE Set $\tvsign \leftarrow \tvsign \: \givmat{i}{j}{\givrot{i,j}}$, $\tvrow \leftarrow \givmat{i}{j}{\givrot{i,j}}\tr \tvrow$.
   \ENDFOR
 \ENDFOR
  \FOR{$j = 1,\dots,\rk$}
   \FOR{$i = \lsd,\dots,\rk+1$}
      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{j,j} / \tvsign_{i,j})$.
      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
      \STATE Set $\tvsign \leftarrow \givmat{i}{j}{\givrot{i,j}} \tvsign$, $\tvcross \leftarrow \tvcross \givmat{i}{j}{\givrot{i,j}}\tr$.
   \ENDFOR
 \ENDFOR
 \STATE Extract $\tvsigndiag$, the first $\rk$ rows of $\tvsign$, and $\tvrot$, the first $\rk$ columns of $\tvcross$.
 \STATE Calculate $\tvfull = \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr$.
 \ENSURE $\tvrot, \tvfull$.
\end{algorithmic}
\caption{Givens decomposition}
\label{alg:givens-decomposition}
\end{algorithm}

@@

In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$. Sampling from each of these should be followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$. Together, this constitutes a collapsed Gibbs move for the target joint posterior distribution \cite{Dyk2008}.

These moves are needed in order to ensure that the sampler can attain any possible values in the parameter space. However, the resulting Markov chain are liable to mix very poorly, since the proposals are dumb random walks. In the following section, we will describe how to draw from the parameter conditional distribution exactly, which greatly accelerates the mixing of the chain.

There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$. For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
%
\begin{IEEEeqnarray}{rCl}
 P(S) & = & (\idmat - S)\inv(I+S)     .
\end{IEEEeqnarray}
%
To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
%
\begin{IEEEeqnarray}{rCl}
 s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
\end{IEEEeqnarray}
%
Use these to construct a skew-symmetric matrix $S$,
%
\begin{IEEEeqnarray}{rCl}
 S_{k,l} & = & \begin{cases}
                s_{k,l}  & k<l \\
                -s_{l,k} & k>l \\
                0        & k=l     ,
               \end{cases}
\end{IEEEeqnarray}
%
and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.

An alternative is to use Givens rotations. First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution. Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ such that,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise,}
                                                 \end{cases}
\end{IEEEeqnarray}
%
and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.

@@

Now introduce,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm \label{eq:transition_matrix_fixed}    ,
\end{IEEEeqnarray}
%
such that the transition matrix may be decomposed into orthogonal components,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \tvrot \lgtmrot + \lgtmorth \label{eq:transition_matrix_decomposition}    .
\end{IEEEeqnarray}
%
Using this, the constraint equation becomes,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
 \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      .
\end{IEEEeqnarray}
%
Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler. In contrast, $\lgtmorth$ and $\tvrot$ are uniquely determined. Sampling from $\postden(\lgtm,\lgtv|\ls{1:\timax})$ thus entails drawing new values from $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.

\subsubsection{Transforming the Prior}
We will need the distribution on $\{\lgtmrot, \tvfull\}$ implied by the prior we chose for $\{\lgtm,\lgtv\}$. These follow from simple properties of the wishart and matrix normal distributions (see \cite{Muirhead1982}). For the matrix normal part,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmrot|\tvrot,\tvfull &\sim& \matrixnormaldist{\tvrot\tr\priormeanmatrix}{\tvfull}{\priorcolumnvariance}     .
\end{IEEEeqnarray}
%
For the inverse Wishart part,
%
\begin{IEEEeqnarray}{rCl}
 \tvfull|\tvrot &\sim& \iwishartdist{\rk}{(\tvrot\tr\priorscalematrix\inv\tvrot)\inv}     .
\end{IEEEeqnarray}
%
Notice that this Inverse Wishart distribution is no longer singular, since $\tvfull$ is $\rk\times\rk$.

@@

@@

There are many ways in which the various Markov kernels could be combined to produce a complete MCMC sampler. Our suggested scheme is shown in algorithm~\ref{alg:complete-degenerate-model-algorithm}. A Markov chain is generated by repeatedly sampling according to this procedure and storing the values returned each iteration. Steps~\ref{step:mh-start} to~\ref{step:mh-end} constitute the collapsed Gibbs move, and steps~\ref{step:gibbs-start} to~\ref{step:gibbs-end} comprise a draw from the constrained parameter conditional.  Each iteration requires a pass of the Kalman filter to calculate the marginal likelihood $\den(\ob{1:\timax}|\lgtm\mcint, \lgtv\mcint)$, plus an additional pass for each Metropolis-Hastings step conducted.

>>>>

More recently, Bayesian approaches have been developed for learning linear Gaussian state space models, in which full posterior distributions for the system parameters are calculated. Since this is not possible analytically (see discussion in \cite{Beal2003}), approximations must be employed. One possibility is to take a variational approach, approximating the full joint posterior distribution as a product of independent marginals \cite{Ghahramani2001,Beal2003,Barber2007}. Alternatively, the true joint posterior can be targeted using Markov chain Monte Carlo (MCMC). This has the advantage of providing consistent estimates not only of the system matrices but also of any function of these quantities (e.g. phase margins, system poles). The price we pay is in computation; it is necessary to allow the Markov chains to run until convergence in order to form good estimates. Generic Metropolis-Hastings strategies such as those adopted in \cite{Ninness2010} are liable to mix very slowly and exact too great a computational demand. However, if conjugate priors are assumed for the various system matrices, then a Gibbs sampler may be implemented which attains substantially improved efficiency and requires almost no algorithm tuning \cite{Wills2012}.

In some systems, the components of the latent state may not be free to vary independently of each other. Such systems are best modelled by a degenerate transition model, in which the transition covariance matrix is singular. This is the case, for example, when the latent state comprises the positions and velocities of multiple points on the same rigid body, meaning that their motion is deterministically coupled. In \cite{Maskell2004,Gustafsson2002}, the authors consider the use of degenerate transition models for various tracking applications. The Gibbs sampler of \cite{Wills2012} is not able to handle degenerate models. In this paper we extend the MCMC learning framework to encompass this scenario. This is achieved by appropriate factorisations of the covariance matrix which allow us still to exploit the fast-mixing Gibbs kernel. In addition, reversible jump moves \cite{Green1995,Green2009} are introduced which enable the sampler to learn explicitly the rank of the transition covariance matrix. The new algorithm is important because it allows us to perform Bayesian learning with a broader class of linear state space models, leading to better descriptions of real systems and improved estimation performance.

@@

A basic linear state space model therefore obeys the following recursive system equations,
%
\begin{IEEEeqnarray}{rCl}
 \ls{\ti} & = & \lgtm \ls{\ti-1} + \tn{\ti} \\
 \ob{\ti} & = & \lgom \ls{\ti}   + \on{\ti}       ,
\end{IEEEeqnarray}
%
where $\ls{\ti} \in \reals^{\lsd}$, $\ob{\ti} \in \reals^{\obd}$. $\lgtm \in \reals^{\lsd\times\lsd}$ and $\lgom \in \reals^{\obd\times\lsd}$ are the transition and observation matrices. $\tn{\ti} \in \reals^{\lsd}$ and $\on{\ti} \in \reals^{\obd}$ are the state disturbance and observation noise variables and we additionally assume that these are drawn from a Gaussian distribution,
%
\begin{IEEEeqnarray}{rCl}
 \tn{\ti} & \sim & \normaldist{0}{\lgtv} \\
 \on{\ti} & \sim & \normaldist{0}{\lgov}     ,
\end{IEEEeqnarray}
%
where $\lgtv$ and $\lgov$ are positive definite covariance matrices. This model can be written equivalently in terms of transition and observation densities,
%
\begin{IEEEeqnarray}{rCl}
 \den(\ls{\ti}|\ls{\ti-1},\lgtm,\lgtv) & = & \normalden{\ls{\ti}}{\lgtm\ls{\ti-1}}{\lgtv} \\
 \den(\ob{\ti}|\ls{\ti},\lgom,\lgov)   & = & \normalden{\ob{\ti}}{\lgom\ls{\ti}}{\lgov}      .
\end{IEEEeqnarray}
%
The initial state $\ls{1}$ may be known or may be assigned a Gaussian prior.

\subsection{Identifiability}
The system is parameterised by four matrices: $\lgtm, \lgom, \lgtv, \lgov$. In some applications, it may be appropriate to treat all of these as completely unknown. However, this scenario can be troublesome for learning algorithms, because the system is not identifiable -- we obtain the same likelihood if the latent state is multiplied by an arbitrary invertible matrix, along with corresponding multiplications of $\lgtm$, $\lgtv$ and $\lgom$. (See e.g. \cite{Roweis1999}.) To resolve the parameter ambiguity, it is necessary to constrain the values of one or more of the system matrices. One option (that taken in \cite{Roweis1999}, for example) is to set $\lgtv$ to the identity matrix. This entails that the nature of the latent states and their relationship to the observations are entirely unknown a priori. In this paper we consider a contrasting approach, where we resolve the ambiguity by specifying the nature of the latent states and setting $\lgom$ to a pre-determined value. This approach has the advantage that the resulting model is more easily interpreted, since we can choose the latent states to be physically meaningful quantities. For example, in the motion capture application studied in section~\ref{sec:simulations}, we choose the latent states to be the positions and velocities of the markers.

@@

The solution we propose here is to use Metropolis-Hastings steps targeting the marginal to ensure irreducibility of the sampler, combined with constrained Gibbs sampling steps, which can only modify a component of the parameters, to accelerate the mixing of the chain. As in the full rank case, the target distribution will be the posterior over parameters and the state trajectory $\postden(\lgtm, \lgtv, \ls{1:\timax})$. This means that each of the Metropolis-Hastings steps targeting the marginal posterior $\postden(\lgtm, \lgtv)$ must be followed by a draw from the state conditional $\postden(\ls{1:\timax}|\lgtm, \lgtv)$ in order to complete the iteration. The two steps together comprise a collapsed Gibbs move \cite{Dyk2008}. Note that the forward-filtering-backward-sampling procedure can be used without modification to draw from the state conditional. See appendix~\ref{app:state-sampling}.

@@

A second decomposition will also be useful. Any positive semi-definite can be factorised in the following way,
%
\begin{IEEEeqnarray}{rCl}
 \lgtv &=& \tvrot \tvfull \tvrot\tr     , \label{eq:givens_decomposition}
\end{IEEEeqnarray}
%
where $\tvfull$ is an $\rk\times\rk$ positive definite matrix and $\tvrot$ is a $\lsd\times\rk$ matrix of orthonormal vectors. In general, this factorisation is not unique. However, here we employ a particular procedure for generating $\tvfull$ and $\tvrot$ which does yield a unique result, ensuring that the transformation from $\lgtv$ to $\{\tvfull, \tvrot\}$ is also bijective. This factorisation is constructed using Givens rotations and thus we refer to it as the Givens decomposition. A Givens rotation matrix has the following structure,
%
\begin{IEEEeqnarray}{rCl}
 \left[\givmat{i}{j}{\givrot{}} - \idmat\right]_{k,l} & = & \begin{cases}
                                                    \cos(\givrot{})-1 & k=l=i \text{ or } k=l=j \\
                                                    \sin(\givrot{}) & k=i,l=j \\
                                                    -\sin(\givrot{}) & k=j,l=i \\
                                                    0 & \text{ otherwise}     ,
                                                 \end{cases} \label{eq:givens_matrix}
\end{IEEEeqnarray}
%
where $\givrot{} \in [-\pi/2,\pi/2]$ is a rotation in the plane of the $i$ and $j$ coordinate directions. The procedure for the Givens decomposition is shown in algorithm~\ref{alg:givens-decomposition}

The Givens decomposition is important because $\tvrot$ is uniquely determined by the state trajectory, whereas $\tvfull$ is free to vary. The computational complexity is $\bigo{\rk\times(\lsd-\rk)}$. (Note that multiplication by a Givens matrix is $\bigo{1}$ due to sparsity.) Details and further discussion is included in appendix~\ref{app:givens-factorisation}.

\begin{algorithm}
\onehalfspacing
\small
\begin{algorithmic}[1]
 \REQUIRE Eigvenvalues and eigenvectors $\tvval,\tvvec$ of positive semi-definite matrix $\lgtv$.
 \STATE Set $\tvcross \leftarrow \idmat[\lsd\times\lsd]$, $\tvrow \leftarrow \idmat[\rk\times\rk]$, $\tvsign \leftarrow \tvvec$.
 \FOR{$i = \rk,\dots,1$}
   \FOR{$j = 1,\dots,i-1$}
      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{i,i})$.
      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
      \STATE Set $\tvsign \leftarrow \tvsign \: \givmat{i}{j}{\givrot{i,j}}$, $\tvrow \leftarrow \givmat{i}{j}{\givrot{i,j}}\tr \tvrow$.
   \ENDFOR
 \ENDFOR
  \FOR{$j = 1,\dots,\rk$}
   \FOR{$i = \lsd,\dots,\rk+1$}
      \STATE Calculate the rotation $\givrot{i,j} = \tan^{-1}(\tvsign_{i,j} / \tvsign_{j,j})$.
      \STATE Construct the Givens matrix $\givmat{i}{j}{\givrot{i,j}}$.
      \STATE Set $\tvsign \leftarrow \givmat{i}{j}{\givrot{i,j}} \tvsign$, $\tvcross \leftarrow \tvcross \givmat{i}{j}{\givrot{i,j}}\tr$.
   \ENDFOR
 \ENDFOR
 \STATE Extract $\tvsigndiag$, the first $\rk$ rows of $\tvsign$, and $\tvrot$, the first $\rk$ columns of $\tvcross$.
 \STATE Calculate $\tvfull = \tvsigndiag \tvrow \tvval \tvrow\tr \tvsigndiag\tr$.
 \ENSURE $\tvrot, \tvfull$.
\end{algorithmic}
\caption{Givens decomposition}
\label{alg:givens-decomposition}
\end{algorithm}

@@

@@

Now introduce,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmorth &=& (\idmat-\tvrot\tvrot\tr) \lgtm \label{eq:transition_matrix_fixed}    ,
\end{IEEEeqnarray}
%
such that the transition matrix may be decomposed into orthogonal components,
%
\begin{IEEEeqnarray}{rCl}
 \lgtm &=& \tvrot \lgtmrot + \lgtmorth \label{eq:transition_matrix_decomposition}    .
\end{IEEEeqnarray}
%
Using this, the constraint equation becomes,
%
\begin{IEEEeqnarray}{rCl}
 \LS{2:\timax} &=& (\tvrot \lgtmrot + \lgtmorth) \LS{1:\timax-1} + \tvrot Z \nonumber \\
 \Rightarrow \LS{2:\timax} &=& \lgtmorth \LS{1:\timax-1} + \tvrot Z' \nonumber      ,
\end{IEEEeqnarray}
%
where $Z'$ is a second arbitrary matrix in the same space as $Z$. Notice that $\lgtmrot$ and $\tvfull$ do not appear in this equation, and thus may be freely altered by the sampler. In contrast, $\lgtmorth$ and $\tvrot$ are uniquely determined. Sampling from $\postden(\lgtm,\lgtv|\ls{1:\timax})$ thus entails drawing new values from $\postden(\lgtmrot,\tvfull|\lgtmorth,\tvrot,\ls{1:\timax})$.

\subsubsection{Transforming the Prior}
We will need the distribution on $\{\lgtmrot, \tvfull\}$ implied by the prior we chose for $\{\lgtm,\lgtv\}$. These follow from simple properties of the Wishart and matrix normal distributions (see \cite{Muirhead1982}). For the matrix normal part,
%
\begin{IEEEeqnarray}{rCl}
 \lgtmrot|\tvrot,\tvfull &\sim& \matrixnormaldist{\tvrot\tr\priormeanmatrix}{\tvfull}{\priorcolumnvariance}     .
\end{IEEEeqnarray}
%
For the inverse Wishart part,
%
\begin{IEEEeqnarray}{rCl}
 \tvfull|\tvrot &\sim& \iwishartdist{\rk}{(\tvrot\tr\priorscalematrix\inv\tvrot)\inv}     .
\end{IEEEeqnarray}
%
Notice that this Inverse Wishart distribution is no longer singular, since $\tvfull$ is $\rk\times\rk$.

@@

In this section we introduce two Metropolis-Hastings kernels which target the marginal posterior $\postden(\lgtm,\lgtv)$. Sampling from each of these should be followed by a draw of the state trajectory from $\postden(\ls{1:\timax}|\lgtm,\lgtv)$. Together, this constitutes a collapsed Gibbs move for the target joint posterior distribution \cite{Dyk2008}. These moves are needed in order to ensure that the sampler can attain any possible values in the parameter space.

There are numerous ways to sample the rotation matrix $\mhrot$ from a suitable proposal distribution $\mhrotppsl$. For example, we could use the Cayley transform \cite{Leon2006}, a bijective mapping from the skew-symmetric matrices to the rotation matrices, defined by,
%
\begin{IEEEeqnarray}{rCl}
 P(S) & = & (\idmat - S)\inv(I+S)     .
\end{IEEEeqnarray}
%
To sample from $\mhrotppsl$, we draw $\half\lsd(\lsd-1)$ independent scalar random variables $\{s_{i,j}\}_{0<i<j<\lsd}$ from any zero-mean distribution; a suitable choice is,
%
\begin{IEEEeqnarray}{rCl}
 s_{k,l} & \sim & \normaldist{0}{\sigma_Q^2} \label{eq:skewsymmetric_proposal}     .
\end{IEEEeqnarray}
%
Use these to construct a skew-symmetric matrix $S$,
%
\begin{IEEEeqnarray}{rCl}
 S_{k,l} & = & \begin{cases}
                s_{k,l}  & k<l \\
                -s_{l,k} & k>l \\
                0        & k=l     ,
               \end{cases}
\end{IEEEeqnarray}
%
and then set $\mhrot=P(S)$. The Cayley transform has the property that $P(-S)=P(S)\inv=P(S)\tr$, which implies that $\mhrotppsl(\mhrot)=\mhrotppsl(\mhrot\tr)$ as required.
An alternative is to use Givens rotations. First sample $i \in [1,\lsd]$, $j \in [1,\lsd]\setminus i$, and $\givrot{} \in [-\pi/2,\pi/2]$ from some zero-mean distribution. Form the Givens matrix $\givmat{i}{j}{\givrot{}}$ according to \eqref{eq:givens_matrix} and use $\mhrot=\givmat{i}{j}{\givrot{}}$. This also has the property that $\givmat{i}{j}{-\givrot{}} = \givmat{i}{j}{\givrot{}}\tr$, implying a symmetric proposal.

@@

There are many ways in which the various Markov kernels could be combined to produce a complete MCMC sampler. Our suggested scheme is shown in algorithm~\ref{alg:complete-degenerate-model-algorithm}. A Markov chain is generated by repeatedly sampling according to this procedure and storing the values returned each iteration. Steps~\ref{step:mh-start} to~\ref{step:mh-end} constitute the collapsed Gibbs move, and steps~\ref{step:gibbs-start} to~\ref{step:gibbs-end} comprise a draw from the constrained parameter conditional.  Each iteration requires two passes of the Kalman filter, in order to calculate the marginal likelihoods $\den(\ob{1:\timax}|\lgtm\mcint, \lgtv\mcint)$ for the Metropolis-Hastings acceptance probability.
